---
title: "MTH361: Probability and Statistics in the Health Sciences"
author: "Probability and Conditional Probability"
date: "February 6/11, 2025"
output: 
  xaringan::moon_reader:
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r echo=FALSE, message=FALSE, warning = FALSE}
library(knitr)

hook_output = knit_hooks$get('output')
knit_hooks$set(output = function(x, options) {
  # this hook is used only when the linewidth option is not NULL
  if (!is.null(n <- options$linewidth)) {
    x = xfun::split_lines(x)
    # any lines wider than n should be wrapped
    if (any(nchar(x) > n)) x = strwrap(x, width = n)
    x = paste(x, collapse = '\n')
  }
  hook_output(x, options)
})

```


### Announcements

+ **Homework 3** 
  - Due **Thursday February 20 at 11:59 pm**
  - Upload to Blueline
  
+ **Lab 3** work day on Thursday February 13
  - Will be due **Friday February 14 at 11:59pm** in Blueline
  
+ **Project Proposal**
  - Due **Tuesday February 25** at 11:59pm in Blueline

<br>
<br>
<br>
<br>

**Book Sections for Next Thursday**: Chapter 3.1 and 3.3

---
class:inverse

<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
.center[
# Probability
]

---

### The "Language" of Probability

**Random Trial**: an experiment or process with two or more possible outcomes, whose occurrence cannot be predicted with certainty

**Event**: the outcome of a random trial that we're most interested in.

**Probability**: the probability of an event is the proportion of times the event would occur if we repeated a random trial many, many times under the exact same conditions.

[Coin Flip Example](https://digitalfirst.bfwpub.com/stats_applet/stats_applet_10_prob.html)

---

### Notation

We use capital letters to describe an event, eg event *E*

We denote the probability of some event *E* as *P(E)*

To be a valid probability, $0 \leq P(E) \leq 1$

For all possible events $A_i$ in a sample space, we have $\sum_iP(A_i) = 1$

---

### Relationship Between Multiple Events

```{r, echo=FALSE}
library(kableExtra)

df <- data.frame(term = c("union", "intersection", "complement"), 
                 notation = c("", "", ''),
                 int = c("", "", ""))

kable(df, col.names = c("Term", "Notation", "Intuitive Terminology"), escape = F)
```

Use venn diagrams to visually show relationships between events! Which is more likely to be true:

- Sarah loves music
- Sarah loves music and she is a data science major

---

### Considering Multiple Events

The way we handle probabilities of two or more events depends on their relationship.

**Mutually Exclusive**: two events *A* and *B* are called mutually exclusive if they cannot both happen at the same time.

- Mutually exclusive events are also called **disjoint**

Examples include: 

- Rolling a 2 or 3 on a dice
- Turning Right and Left
- Flipping a Coin


---

### Considering Multiple Events

**Independent** two events *A* and *B* are called independent if the occurrence of one does not affect the occurence of the other, and vice versa.

- Events that are not independent are **dependent**

Examples include:

- Rolling dice and flipping a coin simultaneously
- Going to college and getting married

---
### Mutually Exclusive or Independent?

The new drug is effective in the trial and the new drug is not effective in the trial

The new drug is effective in the trial and the new drug has side effect

The new drug is effective in the trial and the patient wants to take the drug

<br>
<br>
<br>
Is it possible that two events are both mutually exclusive and independent?
---
### Addition Rule for Mutually Exclusive Events

**Addition Rule**: if two events are mutually exclusive, then 
$$ P(A \cup B) = P(A) + P(B) $$

**Example**: Government data assigns a single cause for each death that occurs in the US. The data show that among people aged 15-24, the probability that a randomly chosen death was accidental is 0.41, 0.16 that it was a homicide, and 0.15 that it was a suicide.

What is the probability that a death is due to one of these three causes?


---
### Complement Rule

**Complement Rule**: the probability that an event does not occur is $$P(A^C) = 1 - P(A)$$


**Example**: Government data assigns a single cause for each death that occurs in the US. The data show that among people aged 15-24, the probability that a randomly chosen death was accidental is 0.41, 0.16 that it was a homicide, and 0.15 that it was a suicide.

What is the probability that a death is due to some other cause?

---
### General Addition Rule

**General Addition Rule**: If two events are not mutually exclusive, then 
$$ P(A \cup B) = P(A) + P(B) - P(A \cap B)$$
**Example**: Insulin injections and oral medication are two key options for treating diabetes. The 2014 National Diabetes Statistics Report provides the breakdown of diabetes treatments among American Adults with diagnosed diabetes. Use the General Addition Rule to find the probability of an adult with diabetes uses either insulin or oral medication.

.pull-right[
```{r, echo=FALSE}
df <- data.frame(term = c("Insulin", "Oral Medication", "Both insulin and oral medication", "Neither insulin or oral medication"), 
                 notation = c(28.7, 71.6, 14.7, 14.4))
kable(df, col.names = c("Treatment", "Percent"), escape = F)
```
]

---
### Multiplication Rule

**Multiplication Rule**: if two events are independent, then 
$$ P(A \cap B) = P(A)*P(B)$$

**Example**: Are insulin injections and oral medication independent for treating diabetes? Explain why or why not.

.pull-right[
```{r, echo=FALSE}
df <- data.frame(term = c("Insulin", "Oral Medication", "Both insulin and oral medication", "Neither insulin or oral medication"), 
                 notation = c(28.7, 71.6, 14.7, 14.4))
kable(df, col.names = c("Treatment", "Percent"), escape = F)
```
]

---
### Example: Doctor Visits

The probability that a visit to a primary care physician’s (PCP) oﬀice results in neither lab work nor referral to a specialist is 40%. Of those coming to a PCP’s oﬀice, 25% are referred to specialists and 45% require lab work

(a). Determine the probability that a visit to the PCP's office results in lab work or referral to a specialist.


(b). Determine the probability that a visit to the PCP's office results in both lab work and referral to a specialist.


---
### Example: Genetics

The gene Prdm9 is thought to regulate hotspots of recombination (crossing over) in mammals. In the people of Han Chinese descent living in the Los Angeles area there are five alleles at the Prdm9 gene: $A_1$, $A_2$, $A_3$, $A_4$, $A_5$. The relative frequencies with which these alleles occur in that population are 0.06, 0.03, 0.84, 0.03, and 0.04, respectively. Is it reasonable to assume that in this population, the two alleles present in any individual are independently sampled from the population as a whole?

What is the probability that an individual has...

(a) Two $A_1$ alleles? $\\$
(b) One $A_1$ allele and one $A_3$ allele $\\$
(c) Does not have two $A_1$ alleles?

---
### Example: Blood Types

The probability that a randomly chosen North American has a particular blood type is listed in the table below.

```{r, echo=FALSE}
df <- data.frame(prob = c(0.374, 0.066, 0.357, 0.063, 0.085, 0.015, 0.034, 0.006))
                 
kable(t(df), col.names = c("O+", "O-", "A+", "A-", "B+", "B-", "AB+", "AB-"), escape = F)
```

1. What is the probability that a randomly selected North American has blood type O?
2. B- blood can be safely donated to anyone who has B+, B-, AB+, or AB- blood. What is the probability that a randomly selected North American can receive blood type B-?
3. Only individuals with O+ blood can receive an O+ blood donation. What is the probability that a randomly selected North American cannot receive blood type O+?
4. What is the probability that a randomly selected North American is Rh+?
5. What is the probability that a randomly selected North American is Rh-?
6. Assume that blood type does not affect the choice of sexual partner. What probability of North American pregnancies will be at risk for Rh incompatibility?

---
class:inverse

<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
.center[
# Conditional Probability
]

---

### Conditional Probability

**Conditional Probability**: the probability of an event occurring, given that some other event has occurred or condition is met.

<br>

.center[**P(A|B)**]

reads as "the probability of **A** given **B**"

Visual Explanation of conditional probability: https://setosa.io/conditional/

---

### Example: Diabetes and CKD

Diabetes and chronic kidney disease (CKD) are two diseases that are increasingly burdening the senior population in the United States. A large national sample of visits to the Veteran's Health Administration (VHA) system indicates that 8.3% of veterans in their 60s have stage 3-5 CKD. This rate is 14.1% among veterans with diabetes and only 6.4% among veterans without diabetes. Express these percents using probability notation. 

---

### Bayes's Theorem

.center[

For any two events A and B, the conditional probability can be written as 

$$ P(A|B) = \frac{P(A \cap B)}{P(B)}$$

]


---
### Example: Diabetes and CKD

A large national sample of visits to the Veteran's Health Administration (VHA) system indicates that 8.3% of veterans in their 60s have stage 3-5 CKD. This rate is 14.1% among veterans with diabetes and only 6.4% among veterans without diabetes. The VHA records also suggest that 25% of veterans in their 60s have diabetes. Find the probability that a randomly selected veteran in their 60s has both diabetes and stage 3-5 CKD.

---
### Extending Bayes

**Law of Total Probability**: Let $A_1$, $A_2$, ... , $A_n$ be disjoint events covering the entire sample space. Then, 
$$P(B) = \sum_{i}P(A_i \cap B) = \sum_{i}P(A_i|B)P(B) = \sum_iP(B|A_i)P(A_i),$$
which gives us Extending Bayes: 
$$ P(A_i|B) =  \frac{P(A_i \cap B)}{P(B)} = \frac{P(A_i \cap B)}{\sum_iP(A_i \cap B)} = \frac{P(B|A_i)P(A_i)}{\sum_jP(B|A_i)P(A_i)}$$


---
### Example: Hand Washing

According to the American Society of Microbiology (2005), the probability that a man washes his hands after using the restroom at an airport is 0.74 and the probability that a woman washes her hands is 0.83. A waiting room in an airport contains 40 men and 60 women. Assume that all passengers-in-waiting are equally likely to use the restroom. What is the probability that the next individual who goes to the restroom will wash their hands?


---
### Additional Helpful Rules

**General Multiplication Rule** = $$ P(A \cap B) = P(A|B)P(B) $$

**Partition** Events = $$ P(A) = P(A \cap B) + P(A \cap B^c) $$


---
### Example: John Hinckley's Trial

In 1981, John Hinckley attempted to assassinate President Ronald Reagan. At this trial, the defense described Hinckley's brain scan as showing human atrophy, which could be evidence of schizophrenia. Approximately 1.5% of Americans suffers from some degree of schizophrenia. Brain atrophy is observed in 30% of schizophrenics, but only 2% of neurotypical individuals. How convincing is the defense's argument?

---

### Example: Rapid HIV Testing

Rapid HIV tests allow for quick diagnosis without expensive laboratory equipment. However, their efficacy has been called into question. In a population of 1517 individuals in Uganda, 4 had HIV but tested negative (False Negative), 166 had HIV and tested positive, 129 did not have HIV but tested positive (false positives) and 1218 did not have HIV and tested negative.

1. What is the probability of a false positive?
2. What is the probability of a false negative?
3. If a randomly sampled individual from this population tests positive on a rapid test, what is the probability that he or she has HIV?

---
### Example: Lupus

Lupus is a medical condition where antibodies that are supposed to attack foreign cells to prevent infections instead see plasma proteins as foreign bodies, leading to a high risk of blood clotting. It is believed that 2% of the population suffer from this disease. The test is 98% accurate if a person has the disease. The test is 74% accurate if a person does not have the disease. There is a line from the Fox television show *House* that is often used after a patient tests positive for lupus: "It's never lupus". Do you this there is truth to this statement? Use appropriate probabilities to support your answer.


---
### Example : Cystic Fibrosis 

Cystic fibrosis (CF) is a hereditary lung disorder that often results in death. It can be inherited only if both parents are carriers of an abnormal gene. In 1989, the CF gene that is abnormal in carriers of cystic fibrosis was identified. The probability that a randomly chosen person carries an abnormal CF gene is 0.03. The CF20m test detects most, but not all harmful mutations of the CF gene. The test is positive for 90% of people who are carriers. It is positive for 1% of people who are not carriers. If you test positive, what is the probability that you are actually a carrier?
