---
title: "MTH361: Probability and Statistics in the Health Sciences"
author: "Hypothesis Testing"
date: "September 26, 2023"
output: 
  xaringan::moon_reader:
    lib_dir: libs
    seal: false
    css: ["default", "metropolis-fonts", "modal.css", "sizeformat.css"]
    includes:
      after_body:
        "js-addins.html"
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: true
---

```{r echo=FALSE, message=FALSE, warning = FALSE}
library(knitr)
library(tidyverse)

hook_output = knit_hooks$get('output')
knit_hooks$set(output = function(x, options) {
  # this hook is used only when the linewidth option is not NULL
  if (!is.null(n <- options$linewidth)) {
    x = xfun::split_lines(x)
    # any lines wider than n should be wrapped
    if (any(nchar(x) > n)) x = strwrap(x, width = n)
    x = paste(x, collapse = '\n')
  }
  hook_output(x, options)
})

```

```{r, include = F, eval = T, cache = T}
clean_file_name <- function(x) {
  basename(x) %>% str_remove("\\..*?$") %>% str_remove_all("[^[A-z0-9_]]")
}
img_modal <- function(src, alt = "", id = clean_file_name(src), other = "") {
  
  other_arg <- paste0("'", as.character(other), "'") %>%
    paste(names(other), ., sep = "=") %>%
    paste(collapse = " ")
  
  js <- glue::glue("<script>
        /* Get the modal*/
          var modal{id} = document.getElementById('modal{id}');
        /* Get the image and insert it inside the modal - use its 'alt' text as a caption*/
        var img{id} = document.getElementById('img{id}');
          var modalImg{id} = document.getElementById('imgmodal{id}');
          var captionText{id} = document.getElementById('caption{id}');
          img{id}.onclick = function(){{
            modal{id}.style.display = 'block';
            modalImg{id}.src = this.src;
            captionText{id}.innerHTML = this.alt;
          }}
          /* When the user clicks on the modalImg, close it*/
          modalImg{id}.onclick = function() {{
            modal{id}.style.display = 'none';
          }}
</script>")
  
  html <- glue::glue(
     " <!-- Trigger the Modal -->
<img id='img{id}' src='{src}' alt='{alt}' {other_arg}>
<!-- The Modal -->
<div id='modal{id}' class='modal'>
  <!-- Modal Content (The Image) -->
  <img class='modal-content' id='imgmodal{id}'>
  <!-- Modal Caption (Image Text) -->
  <div id='caption{id}' class='modal-caption'></div>
</div>
"
  )
  write(js, file = "js-addins.html", append = T)
  return(html)
}
# Clean the file out at the start of the compilation
write("", file = "js-addins.html")
```

class:inverse

<br><br><br>
## MTH361: Probability and Statistics in the Health Sciences
### Hypothesis Tests
#### September 26, 2023


---
### Agenda



---
### Announcements

---
### From Confidence Intervals to Hypothesis Tests

Recall in Lab 5, we have this statement:

*The poll provides strong evidence that the percentage of US adults who suffer from chronic illnesses is below 50%.*

We also know the 95% confidence interval is [0.43, 0.47].

Today we are going to discuss and solve these kind of question in a Hypothesis way: Given a sample, how to make conclusion about the population statistics.

---
### Hypothesis Testing

Hypothesis testing is a method for evaluating the evidence for/against a specific research claim based on a probability, called a p-value. There are four basic steps to this procedure:

- Write the null and alternative hypothesis
- Collect evidence from our sample and calculate a test statistic
- “Translate” the test statistic into a p-value
- Make a conclusion based on the p-value

---
### Step 1: Write Hypotheses

If you want to study something, you need an assumption/expectation/guess for your result $\longrightarrow$ Hypothesis

A Hypothesis contains two mutually exclusive parts:

- Null Hypothesis
- Alternative Hypothesis

Goal: We want to show that our data indicates the probability of null hypothesis happens is quite low (nearly impossible to happen), thus, the alternative hypothesis is more likely to be true.

---
### Step 1: Write Hypotheses

Null Hypothesis $(H_0)$: The null hypothesis represents an assumption or a skeptical statement about the population parameter, and always includes just a single value.

Alternative Hypothesis $(H_A)$: A competing hypothesis contrasts to the null hypothesis, usually is what the original research question hoped to show, prove, or find evidence for.

Note: Alternative hypothesis can be one-sided ("increased", "decreased") or two-sided ("changed", "different").

---
### Step 1: Write Hypotheses

For the following alternative hypotheses: (1) determine whether it is one-sided or two-sided, and (2) state the corresponding null hypothesis

- Patients who take phentermine and topiramate lose weight at a different rate than control patients without these drugs.
- Artesunate is less effective that quinine at treating children with severe malaria.
- Cities that do not have safe-injection sites for drug addicts have higher rates of HIV transmission than cities that do have safe-injection sites.

---
### Step 1: Write Hypotheses

**Example**: Drug maker GlaxoSmithKline (GSK) investigated the potential impact of its oral anti-diabetic drug Avandia on the blood lipid levels of adult diabetics who might benefit from taking Avandia. They were particularly concerned that LDL, "bad" cholesterol, levels might increase. Write the null and alternative hypothesis for their test.

--
$H_0$: LDL will not change after taking Avandia $\\$
$H_A$: LDL will increase after taking Avandia

---
### Step 2: Calculate the Test Statistic 

Given the hypothesis, now we need to use data to show whether the hypothesis is valid or not, through test statistics. 

Test statistic: a number calculated from the sample data that is used to evaluate how far the data we've observed are from the result expected assuming the null hypothesis is true.

Note: We know by now that data does not always reflect the population but we also know that a large enough random sample is pretty close.

---
### Step 2: Calculate the Test Statistic 

Usually the test statistic follows this form:

$$\text{test statistics} = \frac{\text{sample statistics} - \text{expectation}}{\text{standard error}}$$
As we'll see, the choice of test statistic depends on two things

- The type of data we're analyzing
- The assumptions we're willing to make

Different parameters and assumptions will change the test statistics from one hypothesis test to another

---
### Step 3: Calculate the p-value

It's useful to translate the test statistic into another measurement that's consistent for all types of hypothesis tests. 

p-value: the **probability** of observing a sample statistic as or more extreme as the data we have observed, assuming the null hypothesis is true.

**The p-value is not the probability that the null hypothesis is true!**

---
### Step 3: Calculate the p-value

How to covert test statistics to p-value? Recall what we did to convert the Z-score to p-value.

Student t distribution:

- Like the normal distribution, the t-distribution has a symmetric smooth shape.
- The normal distribution assumes that the population standard deviation is known. The t-distribution does not make this assumption.
- The t-distribution is defined by the degrees of freedom. These are related to the sample size.
- The t-distribution is most useful for small sample sizes, when the population standard deviation is no known, or both.
- As the sample size increases, the t-distribution becomes more similar to the normal distribution.

---
### Step 3: Calculate the p-value


---
### Step 4: Make a conclusion





























