---
title: "MTH361: Probability and Statistics in the Health Sciences"
author: "Hypothesis Testing"
date: "September 26, 2023"
output: 
  xaringan::moon_reader:
    lib_dir: libs
    seal: false
    css: ["default", "metropolis-fonts", "modal.css", "sizeformat.css"]
    includes:
      after_body:
        "js-addins.html"
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: true
---

```{r echo=FALSE, message=FALSE, warning = FALSE}
library(knitr)
library(tidyverse)

hook_output = knit_hooks$get('output')
knit_hooks$set(output = function(x, options) {
  # this hook is used only when the linewidth option is not NULL
  if (!is.null(n <- options$linewidth)) {
    x = xfun::split_lines(x)
    # any lines wider than n should be wrapped
    if (any(nchar(x) > n)) x = strwrap(x, width = n)
    x = paste(x, collapse = '\n')
  }
  hook_output(x, options)
})

```

```{r, include = F, eval = T, cache = T}
clean_file_name <- function(x) {
  basename(x) %>% str_remove("\\..*?$") %>% str_remove_all("[^[A-z0-9_]]")
}
img_modal <- function(src, alt = "", id = clean_file_name(src), other = "") {
  
  other_arg <- paste0("'", as.character(other), "'") %>%
    paste(names(other), ., sep = "=") %>%
    paste(collapse = " ")
  
  js <- glue::glue("<script>
        /* Get the modal*/
          var modal{id} = document.getElementById('modal{id}');
        /* Get the image and insert it inside the modal - use its 'alt' text as a caption*/
        var img{id} = document.getElementById('img{id}');
          var modalImg{id} = document.getElementById('imgmodal{id}');
          var captionText{id} = document.getElementById('caption{id}');
          img{id}.onclick = function(){{
            modal{id}.style.display = 'block';
            modalImg{id}.src = this.src;
            captionText{id}.innerHTML = this.alt;
          }}
          /* When the user clicks on the modalImg, close it*/
          modalImg{id}.onclick = function() {{
            modal{id}.style.display = 'none';
          }}
</script>")
  
  html <- glue::glue(
     " <!-- Trigger the Modal -->
<img id='img{id}' src='{src}' alt='{alt}' {other_arg}>
<!-- The Modal -->
<div id='modal{id}' class='modal'>
  <!-- Modal Content (The Image) -->
  <img class='modal-content' id='imgmodal{id}'>
  <!-- Modal Caption (Image Text) -->
  <div id='caption{id}' class='modal-caption'></div>
</div>
"
  )
  write(js, file = "js-addins.html", append = T)
  return(html)
}
# Clean the file out at the start of the compilation
write("", file = "js-addins.html")
```

class:inverse

<br><br><br>
## MTH361: Probability and Statistics in the Health Sciences
### Hypothesis Tests
#### September 26, 2023


---
### Agenda



---
### Announcements

---
### From Confidence Intervals to Hypothesis Tests

Recall in Lab 5, we have this statement:

*The poll provides strong evidence that the percentage of US adults who suffer from chronic illnesses is below 50%.*

We also know the 95% confidence interval is [0.43, 0.47].

Today we are going to discuss and solve these kind of question in a Hypothesis way: Given a sample, how to make conclusion about the population statistics.

---
### Hypothesis Testing

Hypothesis testing is a method for evaluating the evidence for/against a specific research claim based on a probability, called a p-value. There are four basic steps to this procedure:

- Write the null and alternative hypothesis
- Collect evidence from our sample and calculate a test statistic
- “Translate” the test statistic into a p-value
- Make a conclusion based on the p-value

